PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes

### 2023-08-15 manuscript_00.Rmd
resubmission changes

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    keep_md: true
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high-performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encompassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 35 publications. To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six new cone human region markers. CONCLUSION: The PLAE web application provides the eye community a powerful and quick means to test hypotheses related to gene expression across a highly diverse, community-derived database."

keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'svg') # set output device to svg
```

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(glue)
library(tictoc)
#tic()()
#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets [@lunEmptyDropsDistinguishingCells2019a]. We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts. To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019a] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016a] to generate per cell doublet probability scores. To identify cells for removal we used the scores in two ways. First, if an individual cell had a Scrublet score greater than 0.8 (where 1.0 is 100% probability of doublet) and a doubletCell score greater than 1,000,000 (unitless measure where higher value is more likely to be doublet) the cell is labelled as a doublet. Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets. All cutoff values were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.

@@

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` datasets, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021]. We updated the scEiaD database for the PLAE v0.93 web app in six important ways. First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021]. Second we added two ocular outflow tract datasets, one brain choroid plexus, and three cornea datasets to enhance coverage across the eye [@collinSingleCellAtlas2021; @gautamMultispeciesSinglecellTranscriptomic2021; @daniCellularSpatialMap2021; @ligockiMolecularCharacteristicsSpatial2021; @patelMolecularTaxonomyHuman2020; @vanzylCellAtlasAqueous2020]. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons [@heSinglecellTranscriptomeProfiling2020]. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the stringency of the cell retention cutoff for detected unique transcripts per cell from 200 to 300. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20]. As before, the pipeline's scVI integration parameters were chosen rigorously by testing integration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021] (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`).

@@

```{r, fig.height=13, fig.width=11, fig.cap = scEiaD_data_overview_cap, echo = FALSE}

@@

study_count_by_tissue <- meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  #filter(Organ == 'Eye') %>% 
  select(Organ, Tissue,organism, study_accession) %>%
  unique() %>%
  group_by(Organ, Tissue, organism) %>% 
  summarise(Count = n()) %>%
  ungroup() %>% 
  mutate(Organ = factor(Organ, levels = c('Eye','Body'))) %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, label = Count, group = organism)) +
  ylab('Number of Studies') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4) +
  cowplot::theme_cowplot() +
  coord_flip() +
  scale_y_continuous(expand = c(0, 0))  + xlab('') +
  ggforce::facet_col(~Organ, scales = 'free_y', space = 'free')

@@

cell_counts <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
  group_by(CellType_predict, organism) %>%
  summarise(Count = n()) %>% 
  mutate(sCount = log2(Count)) %>% 
  ggplot(aes(x=CellType_predict,y=sCount, label = Count, fill = organism, group = organism)) +
  ylab('log2(Number of Cells)') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4) + 
  cowplot::theme_cowplot() +
  coord_flip() + 
  scale_y_continuous(expand = c(0, 0)) + xlab('')

study_counts_by_cell <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
  group_by(organism, CellType_predict, study_accession) %>%
  summarise(Count = n()) %>% 
  filter(Count > 10) %>% 
  summarise(Count = n()) %>% 
  mutate(Count = case_when(CellType_predict != 'Unlabelled' ~ Count)) %>% 
  mutate(sCount = log2(Count)) %>% 
  ggplot(aes(x=CellType_predict,y=Count, label = Count, fill = organism, group = organism)) +
  ylab('Number of Studies') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4) + 
  cowplot::theme_cowplot() +
  coord_flip() + 
  scale_y_continuous(expand = c(0, 0)) + xlab('')

@@

```{r PLAE_screenshots, fig.height = 10, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
PLAE_screenshots_cap <- fig_cap(name='PLAE_screenshots', caption = 'Screenshots from PLAE app. A. UMAP view showing *Crx* expression with cell type labels, B. Information table showing expression of Crx across celltypes and organism. C. Box plot of selected genes in a user selected cell types across different studies. D. In silico in-situ of Rho. E. Dotplot demonstating efficient display of gene expression across many cell types.')
knitr::include_graphics('fig/figure2 copy.png')
```

@@

```{r echo = F, message = F, }

@@

consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide,
                                           column_labels = consist_diff_organism,
                                           column_split = consist_diff_cts,
                                           row_split = consist_diff_base,
                                           show_row_dend = FALSE,
                                           column_title_rot = 90,
                                           name = 'log2FC', use_raster = TRUE,
                                           row_title_rot = 0) +
  ComplexHeatmap::Heatmap(as.character(query0),
                          row_split = consist_diff_base,
                          col = citation_col,
                          row_labels = consist_diff_genes,
                          show_heatmap_legend = FALSE,
                          column_labels = "Pubmed Gene Citations",
                          column_names_side = 'top') +
  ComplexHeatmap::Heatmap(as.character(query2),
                          row_split = consist_diff_base,
                          col = citation_col,
                          row_labels = consist_diff_genes,
                          name = 'Citation Count',
                          column_labels = "Pubmed Gene + \"Retina\" Citations",
                          column_names_side = 'top')
ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")
queries %>% select(Symbol, Base, `mean log2FC`, `mean padj`, pm_query0, query0, pm_query2, query2) %>% write_csv("supplemental_files/consistently_differentially_expressed_diff_table.csv.gz")
#toc()
```

@@

@@

```{r sup_CT_table,fig.height = 15, fig.width = 20, echo = F, message = F, results = 'asis',fig.cap=sup_CT_table_cap}
# 0
sup_CT_table_cap <- supTab_cap(name = 'sup_CT_table', caption = 'Count and ratio of cell type and predicted cell type labelling assignments split by organisms shows overall strong labelling performance. Cell type to predicted cell type comparisons with a ratio less than 0.01 were filtered out.')
cat("<div id=\"tbl:sup_CT_table\">")
ct_table %>% filter(Ratio > 0.01) %>% flextable() %>% autofit() %>% fit_to_width(7)

>>>>

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    keep_md: true
    fig_caption: yes

bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high-performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encompassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 35 publications. To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six new cone human region markers. CONCLUSION: The PLAE web application provides the eye community a powerful and quick means to test hypotheses related to gene expression across a highly diverse, community-derived database."

keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf', dpi = 1200) # set output device to pdf
```

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(glue)
library(tictoc)
#tic()()
#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
data_dir_vGiga <- '~/data/scEiaD_gigascience/'
data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
#data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
#data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets [@lunEmptyDropsDistinguishingCells2019a]. We then kept cells with the following criteria: less than 10% of the counts were in mitochondrial genes (20% for 10x v3 as this platform has overall higher mitochondrial counts), a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts. To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019a] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016a] to generate per cell doublet probability scores. To identify cells for removal we used the scores in two ways. First, if an individual cell had a Scrublet score greater than 0.8 (where 1.0 is 100% probability of doublet) and a doubletCell score greater than 1,000,000 (unitless measure where higher value is more likely to be doublet) the cell is labelled as a doublet. Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets. All cutoff values were selected by hand inspecting density distributions of the Scrublet and doubletCell scores. Study level distributions of percent mitochondrial genes, total gene counts, and unique genes identified are shown in `r supFig_cap(name = 'qc_metrics', display = 'cite')`.

@@

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` datasets, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021]. We updated the scEiaD database for the PLAE v0.94 web app in six important ways. First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021]. Second we added two ocular outflow tract datasets, one brain choroid plexus, and three cornea datasets to enhance coverage across the eye [@collinSingleCellAtlas2021; @gautamMultispeciesSinglecellTranscriptomic2021; @daniCellularSpatialMap2021; @ligockiMolecularCharacteristicsSpatial2021; @patelMolecularTaxonomyHuman2020; @vanzylCellAtlasAqueous2020]. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons [@heSinglecellTranscriptomeProfiling2020]. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the stringency of the cell retention cutoff for detected unique transcripts per cell from 200 to 300. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20]. As before, the pipeline's scVI integration parameters were chosen rigorously by testing integration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021] (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`).

@@

```{r scEiaD_data_overview, fig.height=13, fig.width=11, fig.cap = scEiaD_data_overview_cap, echo = FALSE}

@@

study_count_by_tissue <- meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  #filter(Organ == 'Eye') %>% 
  select(Organ, Tissue,organism, study_accession) %>%
  unique() %>%
  group_by(Organ, Tissue, organism) %>% 
  summarise(Count = n()) %>%
  ungroup() %>% 
  mutate(Organ = factor(Organ, levels = c('Eye','Body'))) %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, label = Count, group = organism)) +
  ylab('Number of Studies') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4, bg.color = 'white', color = 'black') +
  cowplot::theme_cowplot() +
  coord_flip() +
  scale_y_continuous(expand = c(0, 0))  + xlab('') +
  ggforce::facet_col(~Organ, scales = 'free_y', space = 'free')

@@

cell_counts <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
  group_by(CellType_predict, organism) %>%
  summarise(Count = n()) %>% 
  mutate(sCount = log2(Count)) %>% 
  ggplot(aes(x=CellType_predict,y=sCount, label = Count, fill = organism, group = organism)) +
  ylab('log2(Number of Cells)') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4, bg.color = 'white', color = 'black') + 
  cowplot::theme_cowplot() +
  coord_flip() + 
  scale_y_continuous(expand = c(0, 0)) + xlab('')

study_counts_by_cell <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
  group_by(organism, CellType_predict, study_accession) %>%
  summarise(Count = n()) %>% 
  filter(Count > 10) %>% 
  summarise(Count = n()) %>% 
  mutate(Count = case_when(CellType_predict != 'Unlabelled' ~ Count)) %>% 
  mutate(sCount = log2(Count)) %>% 
  ggplot(aes(x=CellType_predict,y=Count, label = Count, fill = organism, group = organism)) +
  ylab('Number of Studies') +
  geom_bar(stat='identity', position='stack') +
  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4, bg.color = 'white', color = 'black') + 
  cowplot::theme_cowplot() +
  coord_flip() + 
  scale_y_continuous(expand = c(0, 0)) + xlab('')

@@

```{r PLAE_screenshots, fig.height = 10, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
PLAE_screenshots_cap <- fig_cap(name='PLAE_screenshots', caption = 'Screenshots from PLAE app. A. UMAP view showing *Crx* expression with cell type labels, B. In silico in-situ of Cltb, C. Dotplot demonstating efficient display of gene expression across many cell types. D. Box plot of eight genes, with each overlaid point representing an individual study.')
knitr::include_graphics('fig/fig2v02.png')
```

@@

```{r sup_tab_consist_diff, echo = F, message = F }

@@

consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide,
                                           column_labels = consist_diff_organism,
                                           column_split = consist_diff_cts,
                                           row_split = consist_diff_base,
                                           show_row_dend = FALSE,
                                           raster_quality = 10,
                                           column_title_rot = 90,
                                           name = 'log2FC', use_raster = TRUE,
                                           row_title_rot = 0) +
  ComplexHeatmap::Heatmap(as.character(query0),
                          row_split = consist_diff_base,
                          col = citation_col,
                          row_labels = consist_diff_genes,
                          show_heatmap_legend = FALSE,
                          raster_quality = 10,
                          column_labels = "Pubmed Gene Citations",
                          column_names_side = 'top') +
  ComplexHeatmap::Heatmap(as.character(query2),
                          row_split = consist_diff_base,
                          col = citation_col,
                          raster_quality = 10,
                          row_labels = consist_diff_genes,
                          name = 'Citation Count',
                          column_labels = "Pubmed Gene + \"Retina\" Citations",
                          column_names_side = 'top')
ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")
queries %>% select(Symbol, Base, `mean log2FC`, `mean padj`, pm_query0, query0, pm_query2, query2) %>% write_csv("supplemental_files/consistently_differentially_expressed_diff_table.csv.gz")
#toc()
```

@@

```{r quality_control,  fig.height = 10, fig.width = 14, echo = F, message = F, fig.cap=qc_metrics_cap}

qc_metrics_cap <- supFig_cap(name='qc_metrics', caption = 'Study level box plot showing unique genes identified, total gene counts, and percent mitochondrial reads.')

sample_info <- data.table::fread('~/git/scEiaD/data/sample_run_layout_organism_tech_biosample_organ_2022_03_04.tsv')
a <- meta_filter  %>% 
  left_join(sample_info %>% select(sample_accession, Platform) %>% unique()) %>% 
  select(Platform, study_accession, nFeature_RNA) %>% 
  ggplot(aes(y=study_accession, x = nFeature_RNA)) + 
  geom_boxplot() + 
  ggforce::facet_col(~Platform, scales = 'free', space = 'free') +
  cowplot::theme_cowplot()

b <- meta_filter  %>% 
  left_join(sample_info %>% select(sample_accession, Platform) %>% unique()) %>% 
  select(Platform, study_accession, nCount_RNA) %>% 
  ggplot(aes(y=study_accession, x = nCount_RNA)) + 
  geom_boxplot() + 
  ggforce::facet_col(~Platform, scales = 'free', space = 'free') +
  ylab('') +
  cowplot::theme_cowplot() +
  theme(axis.text.y=element_blank())

c <- meta_filter  %>% 
  left_join(sample_info %>% select(sample_accession, Platform) %>% unique()) %>% 
  select(Platform, study_accession, percent_mt) %>% 
  ggplot(aes(y=study_accession, x = percent_mt)) + 
  geom_boxplot() + 
  ggforce::facet_col(~Platform, scales = 'free', space = 'free') +
  ylab('') +
  cowplot::theme_cowplot() +
  theme(axis.text.y=element_blank())

plot_grid(a,b,c, ncol = 3)
```

@@

```{r sup_CT_table,fig.height = 15, fig.width = 20, echo = F, message = F, results = 'asis',fig.cap=sup_CT_table_cap}
# 0
sup_CT_table_cap <- supTab_cap(name = 'sup_CT_table', caption = 'Count and ratio of cell type and predicted cell type labelling assignments split by organisms shows overall strong labelling performance. Cell type to predicted cell type comparisons with a ratio less than 0.01 were filtered out.')
cat("<div id=\"tbl:sup_CT_table\">")
ct_table %>% filter(Ratio > 0.01) %>% flextable() %>% autofit() %>% fit_to_width(12)
